---
title: "Replication of Card and Krueger (1994)"
subtitle: "Minimum Wages and Employment in New Jersey and Pennsylvania"
author: "Peirong"
date: today
---

## Introduction

In this post, I replicate key results from Card and Krueger (1994), *Minimum Wages and Employment: A Case Study of the Fast-Food Industry in New Jersey and Pennsylvania*.  
On April 1, 1992, New Jersey increased its minimum wage from \$4.25 to \$5.05, while neighboring Pennsylvania did not. This setting allows for a difference-in-differences (DiD) research design, where New Jersey stores serve as the treatment group and Pennsylvania stores serve as the comparison group. This approach helps isolate the causal effect of the minimum wage increase on employment outcomes.

The assignment for option 3 asks us to replicate:

1. Some of Table 2.
2. Table 3 (the 9 numbers in rows 1-3 and columns 1-3).
3. Table 4, column (i).

I follow the original data construction rules in the authors' `check.sas` file.

## Main Analysis

### Load Data and Recreate Variables

```{r}
library(dplyr)
library(knitr)
library(ggplot2)

file_path <- "data/public.dat"

col_names <- c(
  "sheet", "chainr", "co_owned", "stater", "southj", "centralj", "northj", "pa1", "pa2", "shore",
  "ncalls", "empft", "emppt", "nmgrs", "wage_st", "inctime", "firstinc", "bonus", "pctaff", "meal",
  "open", "hrsopen", "psoda", "pfry", "pentree", "nregs", "nregs11",
  "type2", "status2", "date2", "ncalls2", "empft2", "emppt2", "nmgrs2", "wage_st2", "inctime2",
  "firstin2", "special2", "meals2", "open2r", "hrsopen2", "psoda2", "pfry2", "pentree2", "nregs2", "nregs112"
)

df <- read.table(file_path, header = FALSE, na.strings = ".", col.names = col_names)

df <- df %>%
  mutate(
    emptot = emppt * 0.5 + empft + nmgrs,
    emptot2 = emppt2 * 0.5 + empft2 + nmgrs2,
    demp = emptot2 - emptot,
    dwage = wage_st2 - wage_st,
    gap = case_when(
      stater == 0 ~ 0,
      wage_st >= 5.05 ~ 0,
      wage_st > 0 ~ (5.05 - wage_st) / wage_st,
      TRUE ~ NA_real_
    ),
    nj = stater,
    bk = as.integer(chainr == 1),
    kfc = as.integer(chainr == 2),
    roys = as.integer(chainr == 3),
    wendys = as.integer(chainr == 4),
    pmeal = psoda + pfry + pentree,
    pmeal2 = psoda2 + pfry2 + pentree2,
    fracft = empft / emptot,
    fracft2 = ifelse(emptot2 > 0, empft2 / emptot2, NA_real_),
    atmin = as.integer(wage_st == 4.25),
    newmin = as.integer(wage_st2 == 5.05),
    closed = as.integer(status2 == 3)
  )

nrow(df)
```

### Replicate Part of Table 2

I replicate several key rows from Table 2: employment and starting wages in wave 1 and wave 2.

```{r}
table2_partial <- data.frame(
  Variable = c(
    "FTE employment (wave 1)",
    "Starting wage (wave 1)",
    "FTE employment (wave 2)",
    "Starting wage (wave 2)",
    "Wage = $5.05 in wave 2 (share)"
  ),
  PA = c(
    mean(df$emptot[df$nj == 0], na.rm = TRUE),
    mean(df$wage_st[df$nj == 0], na.rm = TRUE),
    mean(df$emptot2[df$nj == 0], na.rm = TRUE),
    mean(df$wage_st2[df$nj == 0], na.rm = TRUE),
    mean(df$newmin[df$nj == 0], na.rm = TRUE)
  ),
  NJ = c(
    mean(df$emptot[df$nj == 1], na.rm = TRUE),
    mean(df$wage_st[df$nj == 1], na.rm = TRUE),
    mean(df$emptot2[df$nj == 1], na.rm = TRUE),
    mean(df$wage_st2[df$nj == 1], na.rm = TRUE),
    mean(df$newmin[df$nj == 1], na.rm = TRUE)
  )
) %>%
  mutate(`NJ - PA` = NJ - PA) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

kable(table2_partial, caption = "Partial replication of Table 2")
```

These results closely match those reported in the original paper. For example, wave-1 employment is approximately 20.4 in NJ and 23.3 in PA, while the wave-2 starting wage is about 5.08 in NJ versus 4.62 in PA.

### Replicate Table 3 (Rows 1-3, Columns 1-3)

The assignment asks for the 9 cells from rows 1-3 and columns 1-3 of Table 3:

1. FTE before.
2. FTE after.
3. Change.

```{r}
row1_pa <- mean(df$emptot[df$nj == 0], na.rm = TRUE)
row1_nj <- mean(df$emptot[df$nj == 1], na.rm = TRUE)

row2_pa <- mean(df$emptot2[df$nj == 0], na.rm = TRUE)
row2_nj <- mean(df$emptot2[df$nj == 1], na.rm = TRUE)

row3_pa <- row2_pa - row1_pa
row3_nj <- row2_nj - row1_nj

table3_9_numbers <- data.frame(
  Row = c(
    "1. FTE employment before",
    "2. FTE employment after",
    "3. Change in mean FTE employment"
  ),
  PA = c(row1_pa, row2_pa, row3_pa),
  NJ = c(row1_nj, row2_nj, row3_nj)
) %>%
  mutate(`Difference (NJ - PA)` = NJ - PA) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

kable(table3_9_numbers, caption = "Replication of Table 3 (rows 1-3, columns 1-3)")
```

The difference-in-differences in row 3 is about **2.754**, which is very close to the published value (about 2.76).

### Replicate Table 4, Column (i)

Following the authors' sample restriction in `check.sas`, estimate:

\[
\Delta Employment_i = \alpha + \beta \cdot NJ_i + \varepsilon_i
\]

```{r}
table4_sample <- df %>%
  filter(!is.na(demp), closed == 1 | (closed == 0 & !is.na(dwage)))

model_col1 <- lm(demp ~ nj, data = table4_sample)
coef_table <- summary(model_col1)$coefficients

table4_col1 <- data.frame(
  Term = rownames(coef_table),
  Estimate = coef_table[, "Estimate"],
  `Std. Error` = coef_table[, "Std. Error"],
  `t value` = coef_table[, "t value"],
  `Pr(>|t|)` = coef_table[, "Pr(>|t|)"]
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

kable(table4_col1, caption = "Replication of Table 4, column (i)")
nrow(table4_sample)
```

The coefficient on `nj` is about **2.326** (SE **1.192**), closely matching the paper's column (i) estimate of roughly **2.33 (1.19)**.

## Something Additional

As an additional step, I plot the pre/post mean FTE employment in each state and annotate the DiD estimate.

```{r}
plot_df <- data.frame(
  state = rep(c("Pennsylvania", "New Jersey"), each = 2),
  period = rep(c("Before (Wave 1)", "After (Wave 2)"), times = 2),
  mean_fte = c(row1_pa, row2_pa, row1_nj, row2_nj)
)

did_est <- (row2_nj - row1_nj) - (row2_pa - row1_pa)

ggplot(plot_df, aes(x = period, y = mean_fte, color = state, group = state)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  labs(
    title = "Difference-in-Differences Visualization",
    subtitle = paste0("Estimated DiD effect = ", round(did_est, 3), " FTE workers"),
    x = NULL,
    y = "Average FTE Employment",
    color = "State"
  ) +
  theme_minimal(base_size = 12)
```

The figure makes the result intuitive: Pennsylvania employment declines while New Jersey employment slightly rises, yielding a positive DiD estimate.

## Conclusion

This replication recovers the key findings requested in option 3:

1. **Table 2 (partial):** means closely match the paper.
2. **Table 3 (9 numbers):** replicated successfully, including a DiD of about **2.754**.
3. **Table 4 column (i):** `NJ` coefficient about **2.326 (SE 1.192)**, very close to the published estimate.

Overall, these findings support the paper's main conclusion that the 1992 New Jersey minimum wage increase did not reduce fast-food employment relative to Pennsylvania in this sample, providing evidence against the standard competitive labor market prediction.
